Executive Constraints and Civil Conflict Onset

### 2020-02-11 paper.Rmd
Draft edits

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of conflict. In that time period, upwards to
two-thirds of all countries have experienced at least one minor
insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging
from motivation-based arguments --- such as, for example, the role of
horizontal inequalities [@Cederman2011] --- to opportunity-based
explanations centered on economic factors [@Collier2012; @Collier2009]
as well as the political and structural conditions underlying state
weakness [@Hegre2001; @Fearon2003]. This latter strain of research in
particular has spurred extensive work into examining the
distinguishing features of weak states and whether political
institutions play a significant role in the risk of civil conflict.
Several studies in this vein have argued in favor of institutions
utilizing disaggregated measures to also investigate the specific
types of institutions relevant to the study of civil conflict
[@Taydas2010; @Hegre2015a; @Bartusevic2018].

@@

As mentioned during the theoretical discussion of possible mechanisms,
executive constraints alone are insufficient in determining the
outbreak of conflict. One possible mediator is the relative capacity
of the state. For the present analysis I follow @Bartusevic2018 in
using the Composite Indicator of National Capability (CINC) from the
National Material Capabilities v5.0 dataset (1816-2012) [@Singer1972;
@Singer1987]. The CINC index is aggregated from several demographic,
industrial, and military indicators in order to capture the available
resources a country can bring to bear in the event of a conflict. This
is an imperfect measure that largely misses the state's bureaucratic
and administrative capacity [@Hendrix2010]. [^wbgi] However, given
that the dimension of interest for this analysis is the repressive
capacity of the state, material capabilities, especially military
resources, should still provide an adequate proxy.

@@

```{r reg-stats}
df <- merged.df %>%
    select(rgdpepc, rgdpepc_gro, pop_density, meanelev,
           rlvt_groups_count, neighbour_conflict, ongoing, peace_yrs,
           independence)

@@

```{r fig.cap = "\\label{fig:err} Error plots"}
p1 <- ggplot(nonlg_est, aes(se, interval)) +
    geom_point(col = "#56B4E9", alpha = 0.2, shape = 1) +
    theme_tufte() +
    ylab("Posterior CI Range") +
    xlab("Observed Error")

@@

lth <- filter(merged.df, country_name == "Lithuania", year == 2012) %>%
    select(pop, area_sqkm) %>%
    mutate(area_sqlm = signif(area_sqkm, 3))

india <- filter(merged.df, country_name == "India", year == 1978) %>%
    select(pop, area_sqkm, meanelev) %>%
    mutate(area_sqkm = signif(area_sqkm, 3),
           meanelev = signif(area_sqkm, 3))

@@

Figure \ref{fig:phat} plots the densities of the first 50 posterior
draws for the predicted probabilities of conflict onset for two
groups, country-years that actually experienced an onset of a civil
conflict and country-years that did not. On average, the predicted
probabilities for the actual onset country-years are higher albeit
still extremely low. This can be seen in the summary statistics
reported in table \ref{tab:psum} for the posterior means of the
predicted probabilities. For onset country-years the interquartile
interval is only 0.0485 to 0.209 with a max of 0.556. To better
compare the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.2 cutoff,
which corresponds to the third quartile in the summary table.
`r signif((93 / (259 + 93)) * 100, 3)` percent of onsets and
`r signif((7902 / (7902 + 172)) * 100, 3)` percent of non-onsets are
correctly predicted. The false positive rate is
`r signif((172 / (7902 + 172)) * 100, 3)` percent and the false
negative rate is `r signif((259 / (93 + 259)) * 100, 3)` percent.

@@

```{r fig.width = 7, fig.cap = "\\label{fig:roc} ROC & Precision-Recall Plots"}

@@

>>>>

Since the end of World War II, civil conflict has been the deadliest
and most prevalent form of conflict. In that time period, upwards to
two-thirds of all countries have experienced at least one minor
insurgency and approximately one-third of all countries have
experienced a major civil war with at least 1,000 battle deaths
[@Pettersson2019]. Given the impact of civil conflict and the
increasing availability of conflict related data, the last several
decades has witnessed a veritable explosion in the number of
quantitative studies investigating the causes of civil conflict
[@Blattman2017]. Scholars have proposed numerous theories ranging from
motivation-based arguments --- such as, for example, the role of
horizontal inequalities [@Cederman2011] --- to opportunity-based
explanations centered on economic factors [@Collier2012; @Collier2009]
as well as the political and structural conditions underlying state
weakness [@Hegre2001; @Fearon2003]. This latter strain of research in
particular has spurred extensive work into examining the
distinguishing features of weak states and whether political
institutions play a significant role in the risk of civil conflict.
Several studies in this vein have argued in favor of institutions by
utilizing disaggregated measures to also investigate the specific
types of institutions relevant to the study of civil conflict
[@Taydas2010; @Hegre2015a; @Bartusevic2018].

@@

As mentioned during the theoretical discussion of possible mechanisms,
executive constraints alone are insufficient in determining the
outbreak of conflict. One possible mediator is the relative capacity
of the state. For the present analysis I follow @Bartusevic2018 in
using the Composite Indicator of National Capability (CINC) from the
National Material Capabilities v5.0 dataset (1816-2012)
[@Singer1987]. The CINC index is aggregated from several demographic,
industrial, and military indicators in order to capture the available
resources a country can bring to bear in the event of a conflict. This
is an imperfect measure that largely misses the state's bureaucratic
and administrative capacity [@Hendrix2010]. [^wbgi] However, given
that the dimension of interest for this analysis is the repressive
capacity of the state, material capabilities, especially military
resources, should still provide an adequate proxy.

@@

```{r reg-stats}
df <- merged.df %>%
    select(rgdpepc, rgdpepc_gro, pop_density, meanelev, discrimpop,
           rlvt_groups_count, neighbour_conflict, ongoing, peace_yrs,
           independence)

@@

```{r fig.cap = "\\label{fig:err} Posterior CI and point estimates vs. observed error"}
p1 <- ggplot(nonlg_est, aes(se, interval)) +
    geom_point(col = "#56B4E9", alpha = 0.2, shape = 1) +
    theme_tufte() +
    ylab("Posterior CI Range") +
    xlab("Observed Error")

@@

lth <- filter(merged.df, country_name == "Lithuania", year == 2012) %>%
    select(pop, area_sqkm) %>%
    mutate(area_sqkm = signif(area_sqkm, 3),
           pop = signif(pop, 3))

india <- filter(merged.df, country_name == "India", year == 1978) %>%
    select(pop, area_sqkm, meanelev) %>%
    mutate(pop = signif(pop, 3),
           area_sqkm = signif(area_sqkm, 3),
           meanelev = signif(area_sqkm, 3))

@@

Figure \ref{fig:phat} plots the densities of the first 50 posterior
draws for the predicted probabilities of conflict onset for two
groups, country-years that actually experienced an onset of a civil
conflict and country-years that did not. On average, the predicted
probabilities for the actual onset country-years are higher albeit
still extremely low. This can be seen in the summary statistics
reported in table \ref{tab:psum} for the posterior means of the
predicted probabilities. For onset country-years the interquartile
interval is only 0.0475 to 0.2 with a max of 0.516. To better
compare the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.2 cutoff,
which corresponds to the third quartile in the summary table.
`r signif((93 / (259 + 93)) * 100, 3)` percent of onsets and
`r signif((7902 / (7902 + 172)) * 100, 3)` percent of non-onsets are
correctly predicted. The false positive rate is
`r signif((172 / (7902 + 172)) * 100, 3)` percent and the false
negative rate is `r signif((259 / (93 + 259)) * 100, 3)` percent.

@@

```{r fig.width = 8, fig.cap = "\\label{fig:roc} ROC & Precision-Recall Plots"}

@@

```{r}
models <- setdiff(list.files("./posteriors/"), c("full_model_new", "sim"))

# TODO: this is killing our compile times
all_betas <- mclapply(models, function(f) {
    load(file.path("posteriors", f, "input.RData"))
    beta <- file.path("posteriors", f, "reg_posteriors.csv.gz") %>%
        read_post %>%
        extract("beta")

    colnames(beta)[1:(1 + ncol(X))] <- c("exec_constraints", colnames(X))
    if (interaction_idx > 0)
        colnames(beta)[ncol(beta)] <- paste("exec *", colnames(X)[interaction_idx])

    post_summarise(beta, probs = c(0.1, 0.5, 0.9)) %>%
        mutate(model = f)
}, mc.cores = 4) %>% bind_rows

all_betas %<>%
    rename(codelow = `10%`,
           median = `50%`,
           codehigh = `90%`) %>%
    mutate(model = to_idx(model) %>% as.factor)

desc <- case_when(models == "full_model" ~ "Full Model",
                  models == "full_no_cinc" ~ "Full Model Excl. CINC",
                  models == "full_no_interaction" ~ "Full Model Without Interaction",
                  models == "major_episodes" ~ "Major Conflict Episodes Only",
                  models == "major_no_interaction" ~ "Major Conflict Episodes Without Interaction",
                  models == "unique_no_interaction" ~ "Unique Onsets Without Interaction",
                  models == "unique_onset" ~ "Unique Onsets Only")

df <- data.frame(Model = 1:length(models),
                 Description = desc,
                 stringsAsFactors = F)

kable(df, "latex", align = c("l", "l"), booktabs = T, linesep = "",
      caption = "Model Descriptions") %>%
    kable_styling(latex_options = c("hold_position", "striped"))

```{r fig.width = 7, fig.height = 6}
ggplot(all_betas, aes(x = parameter, y = median, color = model)) +
    geom_point(position = position_dodge(width = 0.8)) +
    geom_errorbar(aes(x = parameter, ymin = codelow, ymax = codehigh),
                  width = 0, size = 1, position = position_dodge(width = 0.8)) +
    geom_hline(yintercept = 0, alpha = 0.5, color = "grey") +
    theme_tufte() +
    scale_color_colorblind() +
    theme(panel.background = element_blank(),
          legend.title = element_blank(),
          axis.line = element_line(color = "black"),
          axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title.x = element_blank()) +
    ylab("Parameter Estimate")
